# Measures of spatial autocorrelation

**Learning objectives:**

- What is spatial autocorrelation
- Tools for measuring spatial autocorrelation
- {rgeoda} package



## Theory

`Tobler’s first law of geography`

> “Everything is related to everything else, but near things are more related than distant things”

Recalling `John Snow's` belief that the cause of cholera was water-borne. 

In summary, `spatial autocorrelation` is the degree to which the values of a variable in one location are correlated with the values of the same variable in neighboring locations.

### Moran's I 

`Moran's Test` $I$ helps us understand if this similarity (nearby locations are more similar to each other than to distant locations) is statistically significant or just due to chance.

It will provide a p-value, if the p-value is low (typically less than 0.05), there is significant spatial autocorrelation in the data, or a non-random pattern.

$$I=\frac{n\sum_{(2)}{w_{ij}z_iz_j}}{S_0\sum_{i=1}^n{z_i^2}}$$

- $z_i=x_i-\bar{x}$

- $\bar{x}=\sum_{i=1}^n{x_i/n}$

- $w_{ij}=\text{spatial weights}$

- $S_0= \sum_{(2)}{w_{ij}}$



```{r}
library(spdep) |> suppressPackageStartupMessages()
# library(parallel)
```



```{r}
glance_htest <- function(ht) c(ht$estimate, 
    "Std deviate" = unname(ht$statistic), 
    "p.value" = unname(ht$p.value))
```

### Case Study: Polish election data

The Polish election data 2015, `pol_pres15`, shows results of elections by type of area.
```{r}
data(pol_pres15, package = "spDataLarge")
pol_pres15 %>%
  dplyr::select(types,I_turnout) %>%
  head
```

```{r message=FALSE,warning=FALSE}
library(tidyverse)
map1 <- ggplot(pol_pres15) +
  geom_sf(aes(geometry=geometry,fill=types),
          linewidth=0.1,
          color="grey80",
          alpha=0.5)+
  scale_fill_viridis_d(option = "C",direction = 1)+
  labs(title="The Polish election data 2015",
       subtitle = "observed",
       fill="Types")+
  ggthemes::theme_map()+
  theme(legend.position = c(-0.2,0))
map1
```

    ?st_make_valid

```{r}
pol_pres15 <- st_make_valid(pol_pres15)
```


```{r}
map1
```


#### Moran's I components



**Construct neighbours list from polygon list**  based on regions with contiguous boundaries.

    ?poly2nb

```{r}
nb_q <- pol_pres15 |> 
  poly2nb(queen = TRUE)
nb_q
```

**Spatial weights for neighbours lists** function adds a weights list with values given by the coding scheme style chosen. B is the basic binary coding.

    ?nb2listw

```{r}
lw_q_B <- nb_q |> 
    nb2listw(style = "B") 
lw_q_B
```

#### Hypothesis testing

For **testing for autocorrelation** we first build a random Normal, `rnorm` distribution of values based on the number of observations.

```{r}
set.seed(1)
x <- pol_pres15 |> 
    nrow() |> 
    rnorm()
```

Then apply the **Moran's I test for spatial autocorrelation** to check the randomness of data.

    ?moran.test
    
$$\left\{\begin{matrix}
 H_0 & p\leq0.05\\ 
 H_a& p>0.05 
\end{matrix}\right. $$


> If the p-value is low (typically less than 0.05), it suggests that there is significant spatial autocorrelation in the data, indicating a non-random pattern.

```{r}
x |> 
    moran.test(lw_q_B, #Spatial weights  
               randomisation = FALSE,
               alternative = "two.sided") 
```

Now see what happens if we use our data and add a gentle trend.

```{r}
library(sf)
coords <- pol_pres15 |> 
    st_geometry() |> # select the geometry
    st_centroid(of_largest_polygon = TRUE)  
```


**Adding a gentle trend** things changed:

```{r}
beta <- 0.0015
t <- coords |> 
    st_coordinates() |> # transform the geometry into lat and long
    subset(select = 1, drop = TRUE) |> 
    (function(x) x/1000)()
```

```{r echo=FALSE,include=FALSE,eval=FALSE}
coords |> 
    st_coordinates()%>%
  subset(select = 1)%>%
  as_tibble()%>%
  mutate(X=X/1000)%>%
  unlist(use.names = F)
```

This is our new data:

     (x + beta * t) 

```{r}
x_t <- (x + beta * t) 

x_t |> 
    moran.test(lw_q_B, 
               randomisation = FALSE,
               alternative = "two.sided") 

```


### Test the residuals

Here we apply the **Moran's I test for residual spatial autocorrelation** to a linear model specification.

    ?lm.morantest
    
```{r}
lm(x_t ~ t) |> 
    lm.morantest(lw_q_B, 
                 alternative = "two.sided") 
```



## Global Measures


> Average level of spatial autocorrelation across all observations


`joincount.test` is similar to Moran's $I$ in `{spdep}`

The **BB join count statistic for k-coloured factors** count test for spatial autocorrelation using a spatial weights matrix in weights list form for testing whether same-colour joins occur more frequently than would be expected if the zones were labelled in a spatially random way.

    ?joincount.test
    ?joincount.multi


```{r}
args(joincount.test)
```

```{r}
args(moran.test)
```



```{r}
Types <- pol_pres15 |> 
        st_drop_geometry() |> 
        subset(select = types, drop = TRUE)
```


```{r}
Types |> 
    table()
```

```{r}
Types |> joincount.multi(listw = lw_q_B) %>%
  head
```

Using an `inverse distance based listw object` releases different results. We first need to identify the **Neighbourhood contiguity by distance** or the neighbours of region points calculated by Euclidean distance in the metric of the points between the lower and the upper bounds.

    ?dnearneigh

- lower distance bound: 0
- upper distance bound: 18300km


```{r}
nb_d183 <- coords |> dnearneigh(0, 18300)
nb_d183
```

Then consider the **Spatial link distance measures** from the neighbour list object (`nb_d183`).

```{r}
gwts <- nb_d183 |> 
    nbdists(coords) |> 
    lapply(function(x) 1/(x/1000))
```

Calculate again the **Spatial weights for neighbours lists** with the `nb2listw` function.

```{r}
lw_d183_idw_B <- nb_d183 |> 
  nb2listw(glist=gwts, style="B")
```



```{r}
Types |> 
  joincount.multi(listw = lw_d183_idw_B) 
```

```{r}
I_turnout <- pol_pres15 |> 
        st_drop_geometry() |> 
        subset(select = I_turnout, drop = TRUE)
```


```{r}
options(scipen = 999)
I_turnout |> 
  moran.test(listw = lw_q_B, randomisation = FALSE) 
```

`lm.morantest` extracts the residuals used for testing to compare with the standard test.
```{r}
lm(I_turnout ~ 1, pol_pres15) |> 
    lm.morantest(listw = lw_q_B) 
```

```{r}
mtr <- I_turnout |> 
    moran.test(listw = lw_q_B) 

mtr |> 
    glance_htest()
```

In the early 1970s, interest was shown in `Monte Carlo tests`, also known as `Hope-type tests` and as `permutation bootstrap`.

**Permutation test for Moran's I statistic** is calculated by using `nsim` random permutations of `x` for the given spatial weighting scheme, to establish the `rank` of the observed statistic in relation to the `nsim` simulated values.

    ?moran.mc
    
```{r}
set.seed(1)
mmc <- I_turnout |> 
    moran.mc(listw = lw_q_B, 
             nsim = 999, 
             return_boot = TRUE) 
```

```{r}
c("Permutation bootstrap" = var(mmc$t), 
  "Analytical randomisation" = unname(mtr$estimate[3]))
```



## Local Measures


Local measures are used to identify clusters or patterns in spatial data at a local level.

- `Local Moran's I` looks at whether a location is part of a cluster of similar values.

- `Local Getis-Ord (Local G)` identifies areas with significantly high or low values compared to their neighbors.

- `Local Geary's C` highlights locations that stand out as being different from nearby areas, whether due to clustering or dispersion.



## The rgeoda package

```{r message=FALSE,warning=FALSE}
library(rgeoda)
Geoda_w <- queen_weights(pol_pres15)
summary(Geoda_w)
```

    ?queen_weights
    
    
**Local Multivariate Geary Statistics**

    ?local_multigeary

```{r}
library(parallel)
lisa <- local_multigeary(Geoda_w, 
    pol_pres15[c("I_turnout", "II_turnout")], 
    cpu_threads = max(detectCores() - 1, 1),
    permutations = 99999, 
    seed = 1)
```



```{r}
all.equal(card(nb_q), 
          lisa_num_nbrs(lisa), 
          check.attributes = FALSE)
```


Let's use the `Local Geary’s` measure to compute local adaptation of Geary's C statistic of spatial autocorrelation.

The `Local Geary` uses squared differences to measure dissimilarity unlike the `Local Moran`. Low values of the Local Geary indicate positive spatial autocorrelation and large refers to negative spatial autocorrelation.


    ?localC_perm returns a pseudo p-value
    
```{r}
lw_q_W <- nb_q |> 
        nb2listw(style = "W")
```
    
    
```{r}
locMvC_p <- pol_pres15 |> 
        st_drop_geometry() |> 
        subset(select = c(I_turnout, II_turnout)) |>
        localC_perm(lw_q_W, nsim=9999, iseed=1) 
```


```{r}
all.equal(lisa_values(lisa), 
          c(locMvC_p),
          check.attributes = FALSE)
```

```{r}
apply(attr(locMvC_p, "pseudo-p")[,c("Pr(z != E(Ci)) Sim", 
                                "Pr(folded) Sim")], 2, range)
```


```{r}
pol_pres15$hs_MvCa <- locMvC_p |> 
  hotspot(Prname = "Pr(folded) Sim",
                    cutoff = 0.0025)
```


```{r}
mvc <- factor(lisa_clusters(lisa), levels=0:2,
              labels = lisa_labels(lisa)[1:3])
```


```{r}
is.na(mvc) <- p.adjust(lisa_pvalues(lisa), "fdr") >= 0.0025
```


```{r}
pol_pres15$geoda_mvc <- droplevels(mvc)
```


```{r}
addmargins(table(spdep = addNA(pol_pres15$hs_MvCa),
                 rgeoda = addNA(pol_pres15$geoda_mvc)))
```


```{r}
library(tmap)

m5 <- tm_shape(pol_pres15) +
    tm_fill("hs_MvCa", 
            palette = RColorBrewer::brewer.pal(4, "Set3")[c(4, 1)],
            colorNA = "grey95", 
            textNA = "Not \"interesting\"",
  title = "Both rounds turnout spdep\nLocal Multivariate Geary C")
```


```{r}
m6 <- tm_shape(pol_pres15) +
    tm_fill("geoda_mvc", 
        palette = RColorBrewer::brewer.pal(4, "Set3")[c(4, 1)],
        colorNA = "grey95", 
        textNA = "Not \"interesting\"",
  title="Both rounds turnout rgeoda\nLocal Multivariate Geary C")
```


```{r}
tmap_arrange(m5, m6, nrow=1)
```


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
